---
title: "Rhizosphere project models"
output:
  html_document:
    fig_caption: yes
---

Load dependencies and set the working directory

```{r message=FALSE, warning=FALSE}

library(phyloseq)
library(phytools)
library(philr)
library(ape)
library(glmnet)
library(ggplot2)
library(ggrepel)
library(dplyr)
# library(broom)
library(broom.mixed)
library(tidyr)
library(reshape2)
library(car)
library(ggfortify)
library(latex2exp)
library(ggtree)

set.seed(139)

# custom functions for data manipulation, exploration and model fitting
source('stat_util.R')
```

### Data preprocessing and primary exploration

Parse the BIOM table, remove rare taxa, process the tree and transform the data using ILR

```{r message=FALSE, warning=FALSE}
# we add pseudo-counts to remove zeros
data <- import_biom(BIOMfilename='stats/table.biom', 
                    treefilename='stats/tree.nwk',
                    parseFunction=parse_taxonomy_greengenes) %>%
    transform_sample_counts(function(x) x + 1)

# binarise the tree and rename internal nodes
phy_tree(data) <- makeNodeLabel(
    multi2di(phy_tree(data)), method='number', prefix='b'
)

# what is the root node
name.balance(phy_tree(data), tax_table(data), 'b1')

# transform all samples (including bare soil samples) for visual inspection
# note: we've disabled weighting (by setting weights to 'uniform'), therefore this is canonical ILR.
transform.wcontrol <- philr(
    t(otu_table(data)), 
    phy_tree(data), 
    part.weights='uniform', 
    ilr.weights='uniform'
)

write.table(transform.wcontrol, file='stats/ilr_all.tsv', sep='\t',
            col.names=NA, quote=FALSE)
```

Explore transformed data through ordination

```{r}
pcoa <- ordinate(data, 'PCoA', distance=dist(transform.wcontrol, method='euclidean'))
(plot.pcoa <- plot_ordination(data, pcoa, color='plant', shape='soil') + 
    geom_point(size=3) + 
    theme(text=element_text(size=11,  family='Helvetica')) +
    theme_bw() + 
    scale_color_brewer(palette='Pastel1'))
dir.create('plots', showWarnings = FALSE)
ggsave(file='plots/pcoa.svg', plot=plot.pcoa, width=7, height=5)
```

### Analyse Soil and Plant effects

We will remove control bulk soil samples and rerun philr transforms

```{r message=FALSE, warning=FALSE}

# subset data
data.wplant <- subset_samples(data, sample_data(data)$plant != 'Control')

# extract observations, tree, taxonomy and metadata from the phyloseq object
observations.wplant <- t(otu_table(data.wplant))
tree.wplant <- phy_tree(data.wplant)
taxonomy.wplant <- tax_table(data.wplant)
meta.wplant <- sample_data(data.wplant)

transform.wplant <- philr(
    observations.wplant, 
    tree.wplant, 
    part.weights='uniform', 
    ilr.weights='uniform'
)

# add metadata to transforms
transform.wplant.wmeta <- as.data.frame(transform.wplant)
transform.wplant.wmeta$soil <- meta.wplant$soil
transform.wplant.wmeta$plant <- meta.wplant$plant
transform.wplant.wmeta$cultivar <- meta.wplant$cultivar
transform.wplant.wmeta$group <- paste(meta.wplant$plant, meta.wplant$soil, sep='@')
```

Select `plant`-separating features

```{r dpi=500}
# cross-validation for l1 lambda parameter; standardise data before running cross-validation
cv.plant <- scale(transform.wplant) %>%
    cv.glmnet(meta.wplant$plant, alpha=1, family='binomial', type.measure='class')

write.table(tidy(cv.plant), file='stats/cv.tsv', sep='\t', quote=FALSE, row.names=FALSE)

(plot.cv <- autoplot(cv.plant) +
        xlab(TeX('$\\log(\\lambda)$')) + 
        theme(text=element_text(size=11,  family='Helvetica')) +
        theme_bw())

ggsave(file='plots/cv.svg', plot=plot.cv, width=8, height=4)
```

```{r}
# extract balances (features) with non-zero coefficients at the highest lambda within 1SE from the CV optimum
features.plant <- select.features(coef(cv.plant, s='lambda.min'))
features.plant
```

Plot feature distributions

```{r}
transform.wplant.long.plant <- transform.wplant %>%
    convert_to_long(get_variable(transform.wplant.wmeta, 'group')) %>%
    filter(coord %in% features.plant)

# consider openning the plot in a separate RStudio tab
ggplot(transform.wplant.long.plant, aes(x=labels, y=value)) +
    geom_boxplot(fill='lightgrey') +
    # scale_y_continuous(trans=signed.log1p.transform) +
    facet_wrap(.~coord, scales='free_x', ncol=5, nrow=8) +
    xlab('Group') + ylab('Balance') +
    theme_bw() + 
    theme(axis.text.x=element_text(angle=90, hjust=1)) + 
    coord_flip()
```

Extract feature metadata (i.e. balance name votes)

```{r}
names.numerator <- extract.balance.names(tree.wplant, taxonomy.wplant, features.plant, 'up', 0.9)
names.denominator <- extract.balance.names(tree.wplant, taxonomy.wplant, features.plant, 'down', 0.9)

# extract balance taxonomy
names.flat.long <- pack.balance.names(names.numerator, names.denominator, paste0('Rank', 4:7))

write.table(names.flat.long, file='stats/balances.tsv', sep='\t',
            col.names=NA, quote=FALSE)
```

Fit mixed models and select models with normally-distributed and homogeneous residuals
```{r}

models.plant <- fit.mixed.models(
    data=transform.wplant.wmeta,
    responses=features.plant,
    'plant  + (1|cultivar)'
)
models.plant.validated <- models.plant[validate.residuals(models.plant, as.factor(transform.wplant.wmeta$plant), 0.2)]
```

We will now aggregate random factor intercept variance along with fixed factor coefficients and remove some unnecessary terms.
```{r}

disambiguate.terms <- function(df) {
    df$term <- ifelse(is.na(df$group), df$term, paste(df$term, df$group, sep='_'))
    df
}

models.plant.summary <- suppressWarnings(
    plyr::ldply(models.plant.validated, tidy, .id = "model") %>%
    disambiguate.terms %>%
    select(-c('effect', 'group', 'std.error', 'statistic')) %>%
    spread(term, estimate)
)

row.names(models.plant.summary) <- strsplit(as.character(models.plant.summary$model), ' ~ ') %>%
    sapply(function(x) x[1])

names(models.plant.summary)
```


```{r}
selected.terms <- c('(Intercept)', 'plantWheat', 'soilDP', 'sd__(Intercept)_cultivar', 'sd__Observation_Residual')
models.plant.summary.clean <- models.plant.summary[selected.terms]

write.table(models.plant.summary.clean, file='stats/models.tsv', sep='\t',
            col.names=NA, quote=FALSE)
```

```{r}
tree.annotation <- data.frame(
    balance=features.plant,
    validated=features.plant %in% row.names(models.plant.summary.clean),
    size=rep(3, length(features.plant))
)

(plot.tree <- ggtree(phy_tree(data)) %<+% tree.annotation +
    geom_nodepoint(aes(color=validated, size=size), alpha=0.5) +
    scale_color_brewer(palette='Pastel1') +
    theme(legend.position='right'))

ggsave(file='plots/tree.svg', plot=plot.tree, width=10, height=8)
```